Modeling and analyzing the Corona-virus warning app with the Isabelle Infrastructure framework

### 2020-07-11 submitted_version_paper.tex
last polish -- is uploaded

\section{Introduction}
\label{sec:intro}
The German Chancellor Angela Merkel has strongly supported the publication of
the mobile phone Corona warning app (CWA; \cite{cwa:github}) by publicly proclaiming that the ``Corona
App deserves your trust'' \cite{bundes:20}. Many millions of mobile phone users
in Germany have downloaded the app with 6 million on the first day.
CWA is one amongst many similar applications that aim at the very important goal
to ``break infection chains'' by providing timely information to users of whether they
have been in close proximity to a person who tested positive for COVID-19.

%The Corona-virus warning app
%has taken a long time to develop being published only on
%16th June 2020. It
%was a quite costly project but this was mainly due to the management of
%Telekom and SAP being in the driving seat. But the app has been designed with great
The app was designed with great
attention on privacy: a distributed architecture \cite{cwa:arch} has been adopted after a long and
heated debate with supporters of a central architecture. The distributed architecture is
based on a very clever application design whereby users' phones broadcast
highly anonymized so called ``Ephemeral IDs'' at physical locations via the
Bluetooth Low Energy Beacon protocol.
The app saves those IDs of people in close proximity.
When at a later date an infected person reports his infection to a central server,
all clients are provided the means to reconstruct the IDs used by this particular
person's device over the last 14 days.
Therefore, any client can evaluate exposure risks locally and stored contact data has never to be shared.
%
%, the unique root ID is published
%and in the daily check all mobile phones connecting to the central server download
%the root IDs of infected people. Since the Ephemeral IDs can be mapped to the root ID
%all Ephemeral IDs that have been saved over the last 14 days allow users' phones to
%regularly check whether their user has been exposed to an infected person and issue
%a warning to the user and recommendation to contact health authorities.
%The warning issued by the Corona warning app entitles to
%having a Corona test done (which at the time of writing is not normally possible).

@@

The app now in use in Germany has been developed based on a protocol proposed by the DP-3T project \cite{dp3t:github}:
A sophisticated security concept
conceived by experts in the field that has strong claims with regard to mathematical support \cite{dp3t:wp} [p2].
%The Corona-virus app has been produced based on a sophisticated security
%concept conceived by experts in the field %to our knowledge, no formal verification
%%
%and has strong claims with regard to mathematical support:
%
%``Such a design builds on strong, mathematically provable support for privacy and data protection goals [...]'' \TODOref{DP-3T Whitepaper p. 2} \smallskip\\
%  \NOTE{DP-3T on their design and how wisely chosen it is ;) We have to use this! Something along the l%ines of:}
%
However, there has been, as of yet, to our knowledge no formal verification been involved.
Even if a ``post-production'' formal specification seems pointless, 
it allows to reveal weak points of the architecture, show that the measures that have
been conceived are suitable to cover those weak points,
or where and to what extend trade offs have to be made due to inherent vulnerabilities.
Given that the protocol implemented by the framework CWA runs on resembles only the most basic
of the three proposed by the DP-3T project,
a formal verified model that stresses the impact or limits of certain security measures
might help lend some more weight to appeals like \cite{dp3t:impl}
%to Apple and Google
to adopt the more sophisticated protocols.
%
%Thereby, we believe
%that our current work is useful to increase the trust in the Corona-virus app necessary for
%its wide adoption which in turn is crucial for it to be efficient.

The contributions of this paper are (a) formal re-engineering of the Corona-virus warning app
(b) providing an additional security and privacy analysis with interactive theorem proving
certification of a novel view on the system architecture including actors, locations, and policies,
(c) a formal definition of a security refinement process that allows to improve a system
based on attacks found by the attack tree analysis and (d) an application of the refinement to
improve security of the Corona-virus app specification.

In this paper, we first provide some background in Section \ref{sec:background}:
we give a detailed overview of the development history %and security and privacy
and relevant parts of the Corona-virus warning app (Section \ref{sec:history})
We then introduce the Isabelle Infrastructure framework (Section \ref{sec:isainf}).
Next, we present our model (Section \ref{sec:model}) and analysis of privacy and
attacks (Section \ref{sec:ana}). The found attack on the first abstract specification
motivates refinement. The formal definition of refinement for the Isabelle Infrastructure
framework is introduced and illustrated on the Corona-virus warning app (Section \ref{sec:ref})
before drawing some conclusions (Section \ref{sec:concl}).

@@

\section{Background and related work}
\label{sec:background}
\subsection{DP-3T and PEPP-PT}
\label{sec:history}
We are mainly concerned with the architecture and protocols proposed by the
DP-3T (\textit{Decentralized Privacy-Preserving Proximity Tracing}) project \cite{dp3t:github}.
The main reason to focus on this particular family of protocols is the
\textit{Exposure Notification Framework} (ENF) \cite{enf:proj}, jointly published by Apple and Google, that adopts
%the main
core principles of the DP-3T proposal. This API is not only used in the German
Corona-virus warning app but has the potential of being widely adopted in future
app developments that might emerge due to the reach of players like Apple and Google.

@@

\subsection{Isabelle Infrastructure framework}
\label{sec:isainf}
%\TODO{Adapt: this is copied from FMBC paper}
The Isabelle Infrastructure is built in the interactive generic theorem prover
Isabelle/HOL \cite{npw:02}. As a framework, it supports formalization and proof of
systems with actors and policies. It originally emerged from verification of insider 
threat scenarios but it soon became clear that the theoretical concepts, like temporal
logic combined with Kripke structures and a generic notion of state transitions were
very suitable to be combined with attack trees into a formal security engineering process
\cite{suc:16} and framework \cite{kam:19a}.
%\subsubsection{Overview}
%\label{sec:isainfover}
Figure \ref{fig:theorystruc} gives an overview of the Isabelle Infrastructure 
framework with its layers of object-logics -- each level below embeds the one
above showing the novel contribution of this paper in blue on the top. 
\begin{figure}
\vspace{-.5cm}
  \begin{center}
  \includegraphics[scale=.35]{theory_structure.pdf}
\end{center}
\vspace{-.5cm}
\caption{Generic Isabelle Infrastructure framework applied to Corona-warning app.}
\label{fig:theorystruc}
\vspace{-.5cm}
\end{figure}

@@

\subsubsection{Instantiation of Framework}
The formal model of the Corona-warning app uses the Isabelle Infrastructure framework instantiating it
by reusing its concept of {\it actors} for users and smartphones whereby locations correspond
to physical locations. The Ephemeral Ids, their sending and change is added to Infrastructures
by slightly adapting the basic state type of infrastructure graphs and accordingly the semantic rules
for the actions move, get, and put. The details of the newly adapted Infrastructure are
presented in Section \ref{sec:model}.
Technically, an Isabelle theory file \texttt{Infrastructure.thy} builds on top of the theories for Kripke 
structures and CTL (\texttt{MC.thy}), attack trees (\texttt{AT.thy}), and security refinement 
(\texttt{Refinement.thy}). Thus all these concepts can be used to specify the formal model
for the Corona-virus warning app, express relevant and interesting properties, and conduct
interactive proofs (with the full support of the powerful and highly automated proof support
of Isabelle).
%The Corona-virus warning app theory itself is an adaptation of the Infrastructure theory of the 
%Isabelle Infrastructure framework and reuses (or slightly adapts) existing concepts. 
%In the remainder of this paper, we introduce the model that we conceived for . 
All Isabelle sources are available online \cite{kam:20gitsc}.

@@

In the Corona-virus warning app, the initial values for the \texttt{igraph} components use
two locations \texttt{pub} and \texttt{shop} to define the following constants (we omit the
data map component \texttt{ex\_locs}).
\begin{ttbox}
ex_loc_ass \ttequiv (\ttlam x. if x = pub then \{''Alice'', ''Bob'', ''Eve''\}  
                 else (if x = shop then \{''Charly'', ''David''\} 
                 else \{\}))
ex_creds \ttequiv (\ttlam x. if x = ''Alice'' then (\{\}, \{\}, Efid 1) else 
              (if x = ''Bob'' then  (\{\},\{\}, Efid 2) else 
              (if x = ''Charly'' then (\{\},\{\}, Efid 3) else
              (if x = ''David'' then (\{\},\{\}, Efid 4) else
              (if x = ''Eve'' then (\{\},\{\}, Efid 5)
               else (\{\},\{\},Efid 0))))))
ex_efids \ttequiv (\ttlam x. if x = pub then \{Efid 1, Efid 2, Efid 5\}
                  else (if x = shop then \{Efid 3, Efid 4\} else \{\}))
ex_knos \ttequiv (\ttlam x. (if x = Actor ''Eve'' then (\ttlam l. \{\} else (\ttlam l. \{\})))
\end{ttbox}  
These components are wrapped up into the following \texttt{igraph}.
\begin{ttbox}
ex_graph \ttequiv
      Lgraph \{(pub, shop)\} ex_loc_ass ex_creds ex_locs ex_efids ex_knos
\end{ttbox}  
Infrastructures are given by the following datatype that 
contains an infrastructure graph of type \texttt{igraph} 
and a policy given by a function that assigns local policies over a graph to
all locations of the graph. 
\begin{ttbox}
{\bf{datatype}} infrastructure = Infrastructure igraph 
                                         [igraph, location] \ttfun policy set
\end{ttbox}
There are projection functions \texttt{graphI} and \texttt{delta} when applied
to an infrastructure return the graph and the local policies, respectively.
The function \texttt{local\_policies} gives the policy for each location \texttt{x}
over an infrastructure graph \texttt{G} as a pair: the first element of this pair is 
a function specifying the actors \texttt{y} that are entitled to perform the actions 
specified in the set which is the second element of that pair.
The local policies definition for the Corona-warning app, simply permits all actions to
all actors in both locations.
\begin{ttbox}
  local_policies G \ttequiv
         (\ttlam x. if x = pub then  \{(\ttlam y. True, \{get,move,put\})\}
          else (if x = shop then \{(\ttlam y. True, \{get,move,put\})\} else \{\}))
\end{ttbox}  
For the Corona-warning app, the initial infrastructure contains the graph \texttt{ex\_graph}
with its two locations pub and shop and is then wrapped up with the local policies
into the infrastructure \texttt{Corona\_scenario} 
that represents the ``initial'' state for the Kripke structure.
\begin{ttbox}
 Corona_scenario \ttequiv Infrastructure  ex_graph local_policies
\end{ttbox}

@@

In addition, we also have a simple implicit version of {\it action refinement}. In an
action refinement, traces may be refined by combining consecutive system events
into atomic events thereby reducing traces.
We can observe this kind of refinement in the second refinement step
%This will be detailed in the second refinement step
of the Corona-virus warning app considered next.

>>>>

\section{Introduction}
\label{sec:intro}
The German Chancellor Angela Merkel has strongly supported the publication of
the mobile phone Corona-virus warning app (CWA; \cite{cwa:github}) by publicly proclaiming that the ``Corona
App deserves your trust'' \cite{bundes:20}. Many millions of mobile phone users
in Germany have downloaded the app with 6 million on the first day.
CWA is one amongst many similar applications that aim at the very important goal
to ``break infection chains'' by providing timely information to users of whether they
have been in close proximity to a person who tested positive for COVID-19.

%The CWA
%has taken a long time to develop being published only on
%16th June 2020. It
%was a quite costly project but this was mainly due to the management of
%Telekom and SAP being in the driving seat. But the app has been designed with great
The app was designed with great
attention on privacy: a distributed architecture \cite{cwa:arch} has been adopted after a long and
heated debate with supporters of a central architecture. The distributed architecture is
based on a very clever application design whereby users' phones broadcast
highly anonymized so called ``Ephemeral IDs'' at physical locations via the
Bluetooth Low Energy Beacon protocol.
The app saves those IDs of people in close proximity.
When at a later date an infected person reports his infection to a central server,
all clients are provided the means to reconstruct the IDs used by this particular
person's device over the last 14 days.
Therefore, any client can evaluate exposure risks locally and stored contact data has never to be shared.
%
%, the unique root ID is published
%and in the daily check all mobile phones connecting to the central server download
%the root IDs of infected people. Since the Ephemeral IDs can be mapped to the root ID
%all Ephemeral IDs that have been saved over the last 14 days allow users' phones to
%regularly check whether their user has been exposed to an infected person and issue
%a warning to the user and recommendation to contact health authorities.
%The warning issued by the Corona warning app entitles to
%having a Corona test done (which at the time of writing is not normally possible).

@@

The app now in use in Germany has been developed based on a protocol proposed by the DP-3T project \cite{dp3t:github}:
A sophisticated security concept
conceived by experts in the field that has strong claims with regard to mathematical support \cite{dp3t:wp} [p2].
%The Corona-virus app has been produced based on a sophisticated security
%concept conceived by experts in the field %to our knowledge, no formal verification
%%
%and has strong claims with regard to mathematical support:
%
%``Such a design builds on strong, mathematically provable support for privacy and data protection goals [...]'' \TODOref{DP-3T Whitepaper p. 2} \smallskip\\
%  \NOTE{DP-3T on their design and how wisely chosen it is ;) We have to use this! Something along the l%ines of:}
%
However, there has, as of yet, to our knowledge no formal verification been involved.
Even if a ``post-production'' formal specification seems pointless, 
it allows to reveal weak points of the architecture, show that the measures that have
been conceived are suitable to cover those weak points,
or where and to what extent trade offs have to be made due to inherent vulnerabilities.
The protocol implemented by the framework CWA resembles only the most basic
of the three proposed by the DP-3T project.
Hence, a formal verified model that stresses the impact or limits of certain security measures
might give more weight to appeals like \cite{dp3t:impl} 
%to Apple and Google
to adopt the more sophisticated protocols.
%
%Thereby, we believe
%that our current work is useful to increase the trust in the Corona-virus app necessary for
%its wide adoption which in turn is crucial for it to be efficient.

The contributions of this paper are (a) formal re-engineering of the CWA
(b) providing an additional security and privacy analysis with interactive theorem proving
certification of a novel view on the system architecture including actors, locations, and policies,
(c) a formal definition of a security refinement process that allows to improve a system
based on attacks found by the attack tree analysis and (d) an application of the refinement to
improve security of the Corona-virus app specification.

In this paper, we first provide some background in Section \ref{sec:background}:
we give a detailed overview of the development history %and security and privacy
and relevant parts of the CWA (Section \ref{sec:history})
We then introduce the Isabelle Infrastructure framework (Section \ref{sec:isainf}).
Next, we present our model (Section \ref{sec:model}) and analysis of privacy and
attacks (Section \ref{sec:ana}). The found attack on the first abstract specification
motivates refinement. The formal definition of refinement for the Isabelle Infrastructure
framework is introduced and illustrated on the CWA (Section \ref{sec:ref})
before drawing some conclusions (Section \ref{sec:concl}).

@@

\section{Background and related work}
\label{sec:background}
\subsection{DP-3T and PEPP-PT}
\label{sec:history}
We are mainly concerned with the architecture and protocols proposed by the
DP-3T (\textit{Decentralized Privacy-Preserving Proximity Tracing}) project \cite{dp3t:github}.
The main reason to focus on this particular family of protocols is the
\textit{Exposure Notification Framework} (ENF) \cite{enf:proj}, jointly published by Apple and Google, that adopts
%the main
core principles of the DP-3T proposal. This API is not only used in the German
CWA but has the potential of being widely adopted in future
app developments that might emerge due to the reach of players like Apple and Google.

@@

\subsection{Isabelle Infrastructure framework}
\label{sec:isainf}
%\TODO{Adapt: this is copied from FMBC paper}
The Isabelle Infrastructure is built in the interactive generic theorem prover
Isabelle/HOL \cite{npw:02}. As a framework, it supports formalization and proof of
systems with actors and policies. It originally emerged from verification of insider 
threat scenarios but it soon became clear that the theoretical concepts, like temporal
logic combined with Kripke structures and a generic notion of state transitions were
very suitable to be combined with attack trees into a formal security engineering process
\cite{suc:16} and framework \cite{kam:19a}.
%\subsubsection{Overview}
%\label{sec:isainfover}
Figure \ref{fig:theorystruc} gives an overview of the Isabelle Infrastructure 
framework with its layers of object-logics -- each level below embeds the one
above showing the novel contribution of this paper in blue on the top. 
\begin{figure}
\vspace{-.5cm}
  \begin{center}
  \includegraphics[scale=.45]{theory_structure.pdf}
\end{center}
\vspace{-.5cm}
\caption{Generic Isabelle Infrastructure framework applied to Corona-warning app.}
\label{fig:theorystruc}
\vspace{-.5cm}
\end{figure}

@@

\subsubsection{Instantiation of Framework}
The formal model of the Corona-warning app uses the Isabelle Infrastructure framework instantiating it
by reusing its concept of {\it actors} for users and smartphones whereby locations correspond
to physical locations. The Ephemeral Ids, their sending and change is added to Infrastructures
by slightly adapting the basic state type of infrastructure graphs and accordingly the semantic rules
for the actions move, get, and put. The details of the newly adapted Infrastructure are
presented in Section \ref{sec:model}.
Technically, an Isabelle theory file \texttt{Infrastructure.thy} builds on top of the theories for Kripke 
structures and CTL (\texttt{MC.thy}), attack trees (\texttt{AT.thy}), and security refinement 
(\texttt{Refinement.thy}). Thus all these concepts can be used to specify the formal model
for the CWA, express relevant and interesting properties, and conduct
interactive proofs (with the full support of the powerful and highly automated proof support
of Isabelle).
%The CWA theory itself is an adaptation of the Infrastructure theory of the 
%Isabelle Infrastructure framework and reuses (or slightly adapts) existing concepts. 
%In the remainder of this paper, we introduce the model that we conceived for . 
All Isabelle sources are available online \cite{kam:20gitsc}.

@@

In the CWA, the initial values for the \texttt{igraph} components use
two locations \texttt{pub} and \texttt{shop} to define the following constants (we omit the
data map component \texttt{ex\_locs}).
\begin{ttbox}
ex_loc_ass \ttequiv (\ttlam x. if x = pub then \{''Alice'', ''Bob'', ''Eve''\}  
                 else (if x = shop then \{''Charly'', ''David''\} 
                 else \{\}))
ex_creds \ttequiv (\ttlam x. if x = ''Alice'' then (\{\}, \{\}, Efid 1) else 
              (if x = ''Bob'' then  (\{\},\{\}, Efid 2) else 
              (if x = ''Charly'' then (\{\},\{\}, Efid 3) else
              (if x = ''David'' then (\{\},\{\}, Efid 4) else
              (if x = ''Eve'' then (\{\},\{\}, Efid 5)
               else (\{\},\{\},Efid 0))))))
ex_efids \ttequiv (\ttlam x. if x = pub then \{Efid 1, Efid 2, Efid 5\}
                  else (if x = shop then \{Efid 3, Efid 4\} else \{\}))
ex_knos \ttequiv (\ttlam x. (if x = Actor ''Eve'' then (\ttlam l. \{\} else (\ttlam l. \{\})))
\end{ttbox}  
These components are wrapped up into the following \texttt{igraph}.
\begin{ttbox}
ex_graph \ttequiv
      Lgraph \{(pub, shop)\} ex_loc_ass ex_creds ex_locs ex_efids ex_knos
\end{ttbox}  
Infrastructures are given by the following datatype that 
contains an infrastructure graph of type \texttt{igraph} 
and a policy given by a function that assigns local policies over a graph to
all locations of the graph. 
\begin{ttbox}
{\bf{datatype}} infrastructure = Infrastructure igraph 
                                         [igraph, location] \ttfun policy set
\end{ttbox}
There are projection functions \texttt{graphI} and \texttt{delta} when applied
to an infrastructure return the graph and the local policies, respectively.
The function \texttt{local\_policies} gives the policy for each location \texttt{x}
over an infrastructure graph \texttt{G} as a pair: the first element of this pair is 
a function specifying the actors \texttt{y} that are entitled to perform the actions 
specified in the set which is the second element of that pair.
The local policies definition for the Corona-warning app, simply permits all actions to
all actors in both locations.
\begin{ttbox}
  local_policies G \ttequiv
         (\ttlam x. if x = pub then  \{(\ttlam y. True, \{get,move,put\})\}
          else (if x = shop then \{(\ttlam y. True, \{get,move,put\})\} else \{\}))
\end{ttbox}  
For the Corona-warning app, the initial infrastructure contains the graph \texttt{ex\_graph}
with its two locations pub and shop and is then wrapped up with the local policies
into the infrastructure \texttt{Corona\_scenario} 
that represents the ``initial'' state for the Kripke structure.
\begin{ttbox}
 Corona_scenario \ttequiv Infrastructure  ex_graph local_policies
\end{ttbox}

@@

In addition, we also have a simple implicit version of {\it action refinement}. In an
action refinement, traces may be refined by combining consecutive system events
into atomic events thereby reducing traces.
We can observe this kind of refinement in the second refinement step
%This will be detailed in the second refinement step
of the CWA considered next.
